Timing and Likelihood of the Origin of Life Derived from Post-Impact Highly Reducing Atmospheres

### 2023-09-12 ms.tex
rest of davids email comments

% Authors
\author{Nicholas F. Wogan}
\affiliation{Department of Earth and Space Sciences, University of Washington, Seattle, WA 98195}
\affiliation{Virtual Planetary Laboratory, University of Washington, Seattle, WA98195}

@@

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history, to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989}. We assume that a life-starting impact is one that produces significant prebiotic nitriles and is not subsequently followed by an ocean-vaporizing impact that destroys any biosphere that has emerged. By considering the fraction of stochastic impact realizations that do not have a life-starting impact, we also estimate the probability of life beginning if Earth history was rerun.

@@

Our estimated timing for biopoiesis is compatible with the earliest well-accepted geologic evidence of life in the form of stromatolites in the 3.5 Ga Pilbara block of western Australia \citep{Walter_1980,Buick_1981,Van_2018}. Older geologic evidence of life exists, such as a $> 3.7$ Ga black shale metamorphosed to graphite with negative $\delta^{13}$C characteristic of biology \citep{Rosing_1999,Ohtomo_2014}. Also, \citet{Bell_2015} discovered graphite inclusions in a 4.1 Ga zircon grain with $^{13}$C depletions compatible with biological activity, but this is a tentative biosignature because there are abiotic mechanisms for making isotopically light carbon \citep{Javaux_2019}. If the \citet{Bell_2015} finding is indeed a sign of life, then it would be generally compatible with our estimated timing in Figure \ref{fig:time_and_mass}, because we find that there is only a $\sim 10\%$ chance of life starting after 4.1 Ga. In general, if life arose from post-impact reducing atmospheres, the origin of life was Hadean, not Archean.

@@

It is temping to extrapolate our results beyond the early Earth to exoplanets, but we must do so with caution because planets orbiting different stars likely have bombardment histories unlike the Hadean. Planets in the habitable zone of a late M-type star during the stellar main-sequence phase were interior to the habitable zone for several hundred million years during the super luminous pre-main-sequence phase \citep{Luger_2015}. \citet{Lichtenberg_2022} use N-body simulations to show that, for planets hosted by late M-dwarfs, large asteroid impacts are not useful for prebiotic chemistry because big impacts most likely occur within 100 million years of planet formation during the stellar pre-main-sequence when the planet is outside of the habitable zone. The pre-main-sequence phase is not an issue for an impact-induced origin of life on planets orbiting sun-like stars because the phase only lasts $\sim 3$ Myrs, while large asteroid impacts occur for 10$^8$ yrs \citep{Lichtenberg_2022}. Therefore, our result that a life-starting impact was likely on the early Earth (92\% chance in an optimistic case) might be best extrapolated to habitable zone exoplanets orbiting sun-like stars.

@@

An additional related caveat is that our calculations assume that an impact that vaporizes the ocean also sterilizes the planet, but this may not be the case if thermophilic microbes survived in the deep ($> 1$ km) subsurface \citep{Sleep_1998}. If ocean-vaporization did not destroy the biosphere, then our results would favor an origin of life earlier than we predict in Figure \ref{fig:time_and_mass}, and a life-starting impact would also be more probable than we have estimated (Figure \ref{fig:probabilities_of_impacts}).

>>>>

% Authors
\author{Nicholas F. Wogan}
\affiliation{Space Science Division, NASA Ames Research Center, Moffett Field, CA 94035}
\affiliation{Virtual Planetary Laboratory, University of Washington, Seattle, WA98195}

@@

Here, we use the \citet{Wogan_2023} results, along with Monte-Carlo simulations of Earth's impact history, to make an alternative estimate for when life most likely emerged during an RNA-first process. Our calculations account for the possibility of planet sterilization by impacts that vaporize the ocean \citep{Sleep_1989,Maher_1988}. We assume that a life-starting impact is one that produces significant prebiotic nitriles and is not subsequently followed by an ocean-vaporizing impact that destroys any biosphere that has emerged. By considering the fraction of stochastic impact realizations that do not have a life-starting impact, we also estimate the probability of life beginning if Earth history was rerun.

@@

Our estimated timing for biopoiesis is compatible with the earliest well-accepted geologic evidence of life in the form of stromatolites in the 3.5 Ga Pilbara block of western Australia \citep{Walter_1980,Buick_1981,Van_2018}. Older geologic evidence of life exists, such as a $> 3.7$ Ga black shale metamorphosed to graphite with negative $\delta^{13}$C characteristic of biology \citep{Rosing_1999,Ohtomo_2014}. Also, \citet{Bell_2015} discovered graphite inclusions in a 4.1 Ga zircon grain with $^{13}$C depletions compatible with biological activity, but this is a tentative biosignature because there are abiotic mechanisms for making isotopically light carbon \citep{Javaux_2019}. If the \citet{Bell_2015} finding is indeed a sign of life, then it would be compatible with our estimated timing in Figure \ref{fig:time_and_mass}, because we find that there is only a $\sim 10\%$ chance of life starting after 4.1 Ga. In general, if life arose from post-impact reducing atmospheres, the origin of life was Hadean, not Archean.

@@

It is temping to extrapolate our results beyond the early Earth to exoplanets, but we must do so with caution because planets orbiting different stars likely have bombardment histories unlike the Hadean. Planets in the habitable zone of a late M-type star during the stellar main-sequence phase were interior to the habitable zone for several hundred million years during the super luminous pre-main-sequence phase \citep{Luger_2015}. \citet{Lichtenberg_2022} use N-body simulations to show that, for planets hosted by late M-dwarfs, large asteroid impacts are not useful for prebiotic chemistry because big impacts most likely occur within 100 million years of planet formation during the stellar pre-main-sequence when the planet is outside of the habitable zone. The pre-main-sequence phase is not an issue for an impact-induced origin of life on planets orbiting sun-like stars because the phase only lasts $\sim 3$ Myrs, while large asteroid impacts occur for $10^8$ yrs \citep{Lichtenberg_2022}. Therefore, our result that a life-starting impact was likely on the early Earth (92\% chance in an optimistic case) might be best extrapolated to habitable zone exoplanets orbiting sun-like stars.

@@

An additional related caveat is that our calculations assume that an impact that vaporizes the ocean also sterilizes the planet, but this may not be the case if thermophilic microbes survived in the deep ($> 1$ km) subsurface \citep{Sleep_1998,Zahnle_2006}. Early life could have also escaped sterilization by being launched to space on rocks during an impact then reseeding the Earth several thousand years later \citep{Wells_2003,Gladman_2005}. If ocean-vaporization did not destroy the biosphere, then our results would favor an origin of life earlier than we predict in Figure \ref{fig:time_and_mass}, and a life-starting impact would also be more probable than we have estimated (Figure \ref{fig:probabilities_of_impacts}).
